### 2020-07-15 paper.Rmd
some work on introduction

---
title             : "An extension of k-means anticlustering to optimize between-group similarity"
shorttitle        : "K-means anticlustering extended"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

Oftentimes, the available data does not represent pairwise 
similarity, but rather numeric, categorical or binary attributes 
[@dry2009]. That is, the data is a two-way table where rows represent 
elements and columns represent variables describing the elements. For 
example, numeric ratings may describe the attributes of word stimuli for 
psycholinguistic experiments. Such data may include Norms for imagery, 
concreteness and orthographic variables [@friendly1982]. In other cases, 
attributes are binary and may represent the presence or absence of 
a feature [@tversky1977]. Attribute data can be converted to pairwise 
dissimilarities for example by computing an appropriate dissimilarity 
measure, such as the pairwise Euclidean or squared Euclidean distance 
across a set of numeric attributes [@brusco2019]. Then, anticlustering 
criteria based on pairwise dissimilarity can be applied to partition the 
elements.

A different approach to anticlustering directly works with feature based 
data and does not require a transformation into pairwise 
dissimilarities: k-means anticlustering. Späth [-@spath1986] and Valev 
[-@valev1998; -@valev1983] independently recognized that by optimizing 
the reversed objective function of the popular k-means algorithm 
[@jain2010; @steinley2006], a data set is partitioned into similar 
groups. They originally---and independently from each other---coined the 
term anticlustering for this purpose. K-means anticlustering maximizes the 
sum of squared Euclidean distances from each element to the center of 
the cluster to which it belongs [cf. @aloise2009np]. As @spath1986 
noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, 
groups only become similar with regard to the mean of the distribution 
of the numeric attributes, which may perform poorly as a measure of 
overall similarity; the spread of the data is not targeted and may 
differ between sets. @papenberg2020 showed that maximizing the 
diversity (based on the pairwise Euclidean distances) is better suited 
to minimize difference with regard to both the mean and the variance of 
the data; it was shown that even an entirely random split usually 
resulted in more similar standard deviations as k-means solely focuses 
on the mean. When requiring that groups are overall similar to each 
other, this focus may be detrimental as similar means can be obtained 
even when the underlying distributions are very different 
[@anscombe1973].

In this paper, I present a simple extension of the k-means objective 
function to maximize simililarity with regard to means and variances 
at the same time.

@@

An example will show that aiming not only for similarity in attribute 
means is not only useful, but also possible. I created 14 data points 
from a normal distribution ($M = 15$, $SD = 3$, see Table 1). The data 
points in Figure 1 illustrate all `r length(partitions)`  ways to 
partition a data set of $N = 14$ into two parts (the Appendix contains 
an explanation how to compute the number of partitions and how to 
generate them). Red point: maximum similarity in means; green point 
= maximum similarity in variance. Blue point: Almost no difference to 
red point wrt similarity in means, but variance is much more similar. 
This blue partition is certainly preferable to the red one when maximal 
overall similarity is required; the red one just optimizes similarity in 
means (which is what maximizing the k-means objective does).

>>>>

---
title             : "A bicriterion extension of k-means anticlustering to optimize between-group similarity"
shorttitle        : "K-means anticlustering extended"

@@

documentclass     : "apa6"
classoption       : "doc"
output            : papaja::apa6_pdf
---

@@

Oftentimes, the available data does not represent pairwise similarity, 
but rather numeric, categorical or binary attributes [@dry2009]. That 
is, the data is a two-way two-mode [@arabie1992] table where rows 
represent the entities of study, and columns represent variables 
describing the elements. For example, numeric ratings may describe the 
attributes of word stimuli for psycholinguistic experiments. Such data 
may include Norms for imagery, concreteness and orthographic variables 
[@friendly1982]. In other cases, attributes are binary and may represent 
the presence or absence of a feature [@tversky1977]. Such data can be 
converted to pairwise dissimilarities for example by computing an 
appropriate dissimilarity measure, such as the pairwise Euclidean or 
squared Euclidean distance across a set of numeric attributes 
[@brusco2019]. Then, anticlustering criteria based on pairwise 
dissimilarity can be applied to partition the elements.

A different approach to anticlustering directly works with feature based 
data and does not require a transformation into pairwise 
dissimilarities. Instead, the sum of the squared Euclidean distances 
between elements and the cluster they are assigned to is minimized 
[brusco2006branch], most prominently using the k-means heuristic 
[@jain2010; @steinley2006]. K-means clustering maximizes between-cluster 
separation and within-cluster homogeneity simultaneously 
[@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed 
k-means objective that is, maximizing the sum of distances between 
elements and cluster centroids -- a data set is partitioned into similar 
groups, thereby reversing the logic of classical clustering. They 
independently from each other coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean 
distances from each element to the center of the cluster to which it 
belongs. As @spath1986 noted, k-means anticlustering thereby directly 
minimizes differences with regard to the means of the numeric variables 
between groups. Hence, groups only become similar with regard to the 
mean of the distribution of the numeric attributes, which may perform 
poorly as a measure of overall similarity; the spread of the data is not 
targeted and may differ between sets. @papenberg2020 showed that 
maximizing the diversity (based on the pairwise Euclidean distances) is 
better suited to minimize difference with regard to both the mean and 
the variance of the data; it was shown that even an entirely random 
split usually resulted in more similar standard deviations as k-means 
solely focuses on the mean. When requiring that groups are overall 
similar to each other, this focus may be detrimental as similar means 
can be obtained even when the underlying distributions are very 
different [@anscombe1973].

In this paper, I present a simple extension of the k-means objective 
function to maximize simililarity with regard to means and variances at 
the same time. The aim is to establish groups that are overall similar 
to each other. It is shown that this is achieved by formulating a 
bicriterion optimization objective for k-means anticlustering. 
Interestingly, under specific---and important---circumstances, this 
reformulation reduces to an augmentation of the original data matrix 
while the optimization objective remains unchanged. This insight has 
important consequences as it allows to adapt standard optimization 
schemes for single criterion anticlustering. I show that the extended 
k-means criterion outperforms other anticlustering criteria when the 
goal is the minimize difference with regard to the mean and variance of 
the different groups. In most cases, both objectives can be optimized
satisfactorily at the same time without any loss on either; when the 
data set is large enough (usually, when $N > XX$), the bicriterion 
optimization leads to > 99.99% objective performance as compared to the 
single optimization of either k-means or k-variance.

@@

An motivating example illustrates that aiming to optimize similarity of 
means of standard deviations in a data set is not only desirable, but 
also expected to be feasible. To this end, I created 14 data points from 
a normal distribution ($M = 15$, $SD = 3$, see Table 1). The data points 
in Figure 1 illustrate all `r length(partitions)`  ways to partition a 
data set of $N = 14$ into two parts. Red point: maximum similarity in 
means; green point = maximum similarity in variance. Blue point: Almost 
no difference to red point wrt similarity in means, but variance is much 
more similar. This blue partition is certainly preferable to the red one 
when maximal overall similarity is required; the red one just optimizes 
similarity in means (which is what maximizing the k-means objective 
does).
